Anxiety: a decision-theoretic perspective

### 2019-02-03 manuscript/manuscript.tex
future unwrapping to be done

Central to many psychiatric disorders is distortion to representations of the
external world and, correspondingly, pronounced changes to learning and decision
making. As a normative framework, decision theory allows us to describe and pose
quantita­tive questions about optimal choice behavior \citep{DayanDaw2008} and is,
therefore, a critical tool for modeling, understanding, and predicting the psychological
and neurobiological mechanisms underlying psychiatric behavior \citep{maia2011, HuysDawDayan2015}.
Recent investigations employing this framework have provided novel and important
insights into psychiatric symptoms and disorders, including anhedonia in major depression \citep{Rutledge2017},
mood instability in bipolar disorder \citep{EldarNiv2015, EldarDolanNiv2016},
habitual behavior in obsessive compulsive and addictive disorders \citep{gillan2016}, and
hallucinations in schizophrenia \citep{powers2017, corlett2018}.

The family of anxiety disorders (which includes generalized anxiety disorder, social
anxiety disorder, panic disorder, and specific phobia) are characterized by excessive
fear and worry about future negative outcomes, and represent the most common
group of psychiatric disorders \citep{kessler2009}. Though the disorders typically
differ in the types of objects or situations that induce fear or anxiety, all share
a set of core disturbances to cognition and behavior. First, the anxiety disorders
involve pessimistic inference, or a tendency to appraise threatening objects or
situations as more likely and/or more severe than they objectively are. Clinical
anxiety is also associated with second-order conditioning, or learning to fear
neutral cues indirectly associated with threat cues. Finally, anxiety is associated
with sustained avoidance, or the tendency for individuals to flee or avoid threatening
situations. As an example, an individual is mugged in an alleyway. In a pathological
response, this individual may exhibit persistent worry that it will happen again
(pessimistic inference), increased anxiety towards the street or time of day
at which it occurred (second-order conditioning), and increased effort to stay
away from any cue which reminds them of the event (avoidance).

Several studies have attempted to provide some insight into why some of this
may be in anxiety disorders \cite{Maia2010, norbury2018}, though no account has
tied these symptoms together. For example, \citep{Maia2010} provided a simple
account of how simple actor-critic reinforcement learning models could account
for sustained avoidance in rodent shuttle-box experiments \citep{servatius2008}.
As has been discussed elsewhere \citep{Moutoussis2017}, such an account fails to account for why avoidance
is maintained even with non-threatening outcomes and do not account for why exposure
therapy only works for some individuals. Other accounts have attempted to explain
pessimistic inference (cite learning rate), but do not mention where from.

In the remainder of this article, we present a simple decision theoretic account
of how the core symptoms of anxiety may arise and be maintained. Specifically, we
demonstrate that the loss of optimistic inference in the form of reduced expectations
of future reward can yield negatively biased estimates of value in sequential
decision environments. In such an event, pessimistic inference, second-order
conditioning, and chronic avoidance naturally follow.

@@

These sorts of problems have been explored in the safe reinforcement learning
and robust control literature \citep{Garcia2015}. There, we are concerned
with programming agents that identify the optimal policy under pessimistic
assumptions. A number of approaches have been proposed in that literature. Here,
we focus on one particular simple implemented proposed by \cite{Gaskett2003}, namely
*w-pessimism*:

@@

In the following sections, we present the results of simulations of pessimistic
learning in a variety of MDP environments. In all simulations, the state-action
values under the optimal policy, $Q_*(s,a)$, were solved for directly through
Q-value iteration \citep{SuttonBarto1998,SuttonBarto1998,bertsekas2005}. The
details of this algorithm are presented in the appendix. Model parameters are
reported for each environment in their corresponding sections. All simulations
were carried out with the python programming language and are publicly available
on Github (https://github.com/szorowi1/SecretFunTimes).

@@

In addition to pessimistic inference, all anxiety disorders share pathological
avoidance as core symptom \citep{dsm5, Krypotos2015, Arnaudova2017}. Avoidance
behaviors seekto increase or maintain the temporal, physical and/or mental distance
between an individual and a threatening situation. Although avoidance is a natural response to danger, excessive avoidance as is observed in anxiety disorders can be highly disruptive to everyday
functioning \cite{Salter2004} and frequently function so as maintain anxiety
by preventing extinction \citep{Arnaudova2017}. For example, an individual with social anxiety may avoid
attending large social events for fear of public embarrassment only to never learn
that such worry is not justified.

@@

This free choice premium has been demonstrated multiple times in humans \citep{Suzuki1997,Leotti2011,Leotti2014,Cockburn2014} using a variety of decision
paradigms. For the present purposes, we will describe the experiments presented in
\citep{Leotti2011,Leotti2014}. In these experiments, participants complete a
two-stage decision making task (Figure 5). In the first stage, participants are
make a choice between two cues: one leading to free choice and the other leading
to a fixed choice. In the second stage, participants are able to choose between
a second set of cues (free choice) or are forced to choose a cue (fixed choice).
Unbeknownst to participants, all cues yield reward following the same outcome
distribution; thus, no cue is advantageous in the long-run. The free choice
premium is measured as a preference for the cue leading to free choice. \cite{Leotti2011,Leotti2014}
find participants still prefer the free choice cue, despite this conferring no
objective benefit to participants.

@@

All of the anxiety disorders share core symptoms including pessimistic inference,
second-order conditioning, and sustained avoidance. In this article, we offered
a simple computational account of how these aberrations may arise and be maintained.
Following the literature in Markov decision processes, we showed how a simple
breakdown in expectations about the likelihood of encountering future negative
outcomes (as a result of an unpredictable environment or unreliable policy) can
effectively backpropagate negative expectations through an agent's internal value
map. The result of which yields expectations and behaviors resembling that observed
in pathological anxiety. Importantly, we also showed that such a model can account
for aberrant behavior previously observed in individuals with anxiety in several
assays of psychiatric behavior. By no means a complete theory of anxiety, this
account helps to at least tie together several core symptoms of anxiety in one
account.

@@

For model-based decision making, one possibility is biased planning. In large
state spaces, online sequential decision making suffers from the curve of
dimensionality: in planning a series of actions to take, comparing between
multiple options rapidly becomes computationally intractable when the number of
options and depth of search becomes even moderately large. Thus, heuristics for
reducing the size of this search can be useful. Biased pruning of search paths
(such as immediately discarding paths that require some minor loss in pursuit of
larger gains) may result in maladaptive decision making (Huys et al., 2012).
Alternately, if planning relies on internal sampling from previously experienced
episodes in order to make predictions about future value, then biased sampling
may similarly result in maladaptive decision making. Recent work has shown how
finite sampling during planning can result in risk aversion and the overrepresentation
of low probability but strongly negative outcomes (Lieder et al., 2018). Such a
mechanism is prima facie in line with results showing the availability in memory
of negative outcomes being higher for patients with anxiety (Borkovec et al., 1999;
Miranda and Mennin, 2007).

Within the realm of model-free decision making, several studies have found evidence
suggesting increased learning rates for negative reward prediction errors in
individuals with elevated levels of anxiety \citep{Aylward, Huang2017, Harle2017
garrett2018}. Though an asymmetry in sensitivity to
positive and negative reward prediction errors bears superficial similarity, for
the reasons described above merely attending more to negative information does not
necessarily predict increased avoidance insofar that avoidance can successfully
isolate negative outcomes. One possibility, and topic of future research, is how
prior beliefs about the reward statistics of the current environment dictate
learning rates. A second possibility is prioritized replay (Russek et al., 2017;
Mattar and Daw, 2018). Offline replay of previous experiences (through hippocampal
mechanisms) are known to facilitate learning. It has been proposed that biased
replay could bias value estimates (Gagne et al., 2018). This is another exciting
area for future research.

>>>>

Many psychiatric disorders involve distortions to representations of the external world and, as a likely consequence, pronounced changes to learning and decision making. As a normative framework, decision theory allows us to describe and pose quantita­tive questions about optimal choice behavior \citep{DayanDaw2008} and is, therefore, a critical tool for modeling, understanding, and predicting the psychological and neurobiological mechanisms underlying psychiatric behavior \citep{maia2011, HuysDawDayan2015}. Recent investigations employing this framework have provided novel and important insights into psychiatric symptoms and disorders, including anhedonia in major depression \citep{Rutledge2017}, mood instability in bipolar disorder \citep{EldarNiv2015, EldarDolanNiv2016}, habitual behavior in obsessive compulsive and addictive disorders \citep{gillan2016}, and hallucinations in schizophrenia \citep{powers2017, corlett2018}.

The family of anxiety disorders (which includes generalized anxiety disorder, social anxiety disorder, panic disorder, and specific phobia) are characterized by excessive fear and worry about future negative outcomes, and represent the most common group of psychiatric disorders \citep{kessler2009}. Though the disorders typically differ in the types of objects or situations that induce fear or anxiety, all share a set of core disturbances to cognition and behavior. First, the anxiety disorders involve pessimistic inference, or a tendency to appraise threatening objects or situations as more likely and/or more severe than they objectively are. Clinical anxiety is also associated with second-order conditioning, or learning to fear neutral cues indirectly associated with threat cues. Finally, anxiety is associated with avoidance, or the tendency to flee or avoid threatening situations. As an example, an individual with a clinical phobia of dogs may show chronic worry about coming into contact with aggressive dogs (pessimistic inference), may develop fear towards parks or other public spaces where they may encounter dogs (second-order conditioning), and might exert greater effort to avoid contact with dogs or locations associated with dogs (avoidance).

Recent work has begun to shine some light into the mechanisms underlying these symptoms. For example, a number of studies have found that anxiety is correlated with increased learning from worse-than-predicted outcomes \citep{Harle2017, Garrett2018, Aylward2019}. Systematic biases in learning could result in pessimistic inference by virtue of negatively biasing those same cached values used to make predictions at the time of retrieval. This begs the question as to why learning from negative prediction errors is greater in anxiety, and others have suggested this may reflect prior expectations in anxiety; that pessimistic expectations bias learning from expectation-congruent information. This raises the possibility of circular arguments.

Other studies have attempted to provide novel insights into avoidance. \cite{Maia2010} offered a reinforcement learning-based account of maintained avoidance in rodent conditioning tasks, an empirical finding that previous theories of avoidance could not account for \citep{Krypotos2015}. They argue that sustained avoidance is possible insofar that a well-established avoidance response generates no reward prediction error, such that no new learning takes place, and avoidance continues. As is discussed elsewhere \citep{Moutoussis2017}, the framework cannot account for sustained avoidance when an agent is accidentally exposed to an extinguished outcome. Returning to the previous example, the framework cannot account for why dog phobia remains for an individual in spite of an unplanned exposure to a friend's benign dog. More broadly, the framework cannot for why avoidance, and anxiety in general, is so persistent even with exposure therapy \citep{Craske2014}.

In the remainder of this article, we present a simple decision theoretic account of how the core symptoms of anxiety may arise and be maintained. Specifically, we demonstrate that how a loss of confidence in the reliability of the self or the environment can result in the catastrophic spread of negative value in sequential decision environments. We show how, under these assumptions, pessimistic inference, second-order conditioning, and chronic avoidance naturally follow, as well as a number of other behaviors previously found to be correlated with anxiety.

@@

These sorts of problems have been explored in the safe reinforcement learning and robust control literature \citep{Garcia2015}. There, we are concerned with programming agents that identify the optimal policy under pessimistic assumptions. A number of approaches have been proposed in that literature. Here, we focus on one particular simple implemented proposed by \cite{Gaskett2003}:

@@

In the following sections, we present the results of simulations of pessimistic learning in a variety of MDP environments. In all simulations, the state-action values under the optimal policy, $Q_*(s,a)$, were solved for directly through Q-value iteration \citep{SuttonBarto1998, SuttonBarto1998, bertsekas2005}. The details of this algorithm are presented in the appendix. Model parameters are reported for each environment in their corresponding sections. All simulations were carried out with the python programming language and are publicly available on Github (https://github.com/szorowi1/SecretFunTimes).

@@

In addition to pessimistic inference, all anxiety disorders share pathological avoidance as core symptom \citep{dsm5, Krypotos2015, Arnaudova2017}. Avoidance behaviors seek to increase or maintain the temporal, physical and/or mental distance between an individual and a threatening situation. Although avoidance is a natural response to danger, excessive avoidance as is observed in anxiety disorders can be highly disruptive to everyday functioning \citep{Salter2004} and frequently function so as maintain anxiety by preventing extinction \citep{Arnaudova2017}. For example, an individual with social anxiety may avoid attending large social events for fear of public embarrassment only to never learn that such worry is not justified.

@@

This free choice premium has been demonstrated multiple times in humans \citep{Suzuki1997, Leotti2011, Leotti2014, Cockburn2014} using a variety of decision paradigms. For the present purposes, we will describe the experiments presented in \citep{Leotti2011, Leotti2014}. In these experiments, participants complete a two-stage decision making task (Figure 5). In the first stage, participants are make a choice between two cues: one leading to free choice and the other leading to a fixed choice. In the second stage, participants are able to choose between a second set of cues (free choice) or are forced to choose a cue (fixed choice). Unbeknownst to participants, all cues yield reward following the same outcome distribution; thus, no cue is advantageous in the long-run. The free choice premium is measured as a preference for the cue leading to free choice. \cite{Leotti2011, Leotti2014} find participants still prefer the free choice cue, despite this conferring no objective benefit to participants.

@@

All of the anxiety disorders share core symptoms including pessimistic inference, second-order conditioning, and sustained avoidance. In this article, we offered a simple computational account of how these aberrations may arise and be maintained. Following the literature in Markov decision processes, we showed how a simple breakdown in expectations about the likelihood of encountering future negative outcomes (as a result of an unpredictable environment or unreliable policy) can effectively backpropagate negative expectations through an agent's internal value map. The result of which yields expectations and behaviors resembling that observed in pathological anxiety. Importantly, we also showed that such a model can accountfor aberrant behavior previously observed in individuals with anxiety in several assays of psychiatric behavior. By no means a complete theory of anxiety, this account helps to at least tie together several core symptoms of anxiety in one account.

@@

For model-based decision making, one possibility is biased planning. In large state spaces, online sequential decision making suffers from the curve of dimensionality: in planning a series of actions to take, comparing between multiple options rapidly becomes computationally intractable when the number of options and depth of search becomes even moderately large. Thus, heuristics for reducing the size of this search can be useful. Biased pruning of search paths (such as immediately discarding paths that require some minor loss in pursuit of larger gains) may result in maladaptive decision making \citep{Huys2012}. Alternately, if planning relies on internal sampling from previously experienced episodes in order to make predictions about future value, then biased sampling may similarly result in maladaptive decision making. Recent work has shown how finite sampling during planning can result in risk aversion and the overrepresentation of low probability but strongly negative outcomes (Lieder et al., 2018). Such a mechanism is prima facie in line with results showing the availability in memory of negative outcomes being higher for patients with anxiety \citep{Borkovec1999, Miranda2007}.

Within the realm of model-free decision making, several studies have found evidence suggesting increased learning rates for negative reward prediction errors in individuals with elevated levels of anxiety \citep{Harle2017, Garrett2018}. Though an asymmetry in sensitivity to positive and negative reward prediction errors bears superficial similarity, for the reasons described above merely attending more to negative information does not necessarily predict increased avoidance insofar that avoidance can successfully isolate negative outcomes. One possibility, and topic of future research, is how prior beliefs about the reward statistics of the current environment dictate learning rates. A second possibility is prioritized replay (Russek et al., 2017; Mattar and Daw, 2018). Offline replay of previous experiences (through hippocampal mechanisms) are known to facilitate learning. It has been proposed that biased replay could bias value estimates (Gagne et al., 2018). This is another exciting area for future research.
